---
title: "R Notebook"
output: html_notebook
---

## Libraries management 
```{r}
if( !require(pacman)) {
  install.packages("pacman")
  library(pacman)
}

p_load(tidyverse)
p_load(iheatmapr)
p_load(tidyselect)

base_dir <- here::here() # "/home/ignatius/PostDoc/2019/Sepsis"
source( file.path( base_dir, "Source/Common/helper_functions.R") )
source( file.path( base_dir, "Source/Common/common_parameters.R") )

```

nohup time Rscript --vanilla -e "rmarkdown::render('plot_perc_kegg_pathway_mapped.Rmd', output_file='plot_perc_kegg_pathway_mapped.html') " > plot_perc_kegg_pathway_mapped.log 2>&1 &


## Global Parameters
```{r}
best_e_value_to_use <- 10^-3

## Parameters to select rows for printing the heat maps
## Show only rows with >= 30% completeness for > 80% of the strains
percent_kegg_nodes_mapped_threshold <- 30
proportion_of_strains_above_threshold <- 0.8 
```

## Directories Management
```{r}

kegg_annotations_dir <- file.path( owncloud_dir, "Multi-omics/Results/KEGG_Annotations")

kegg_figures_dir <- file.path( oc_results_dir, "KEGG_Annotations/Figures")

create_dir_if_not_exists( file.path( oc_results_dir, "KEGG_Annotations/Figures"))

kegg_annotations_e_value_dir <- file.path( oc_results_dir, "KEGG_Annotations", paste( "E_value_", best_e_value_to_use, sep=""), "Map_to_Uniprot" )

kegg_ko_tables_dir <- file.path( kegg_annotations_dir, "KEGG_ko_tables") 

```

## Read in the KEGG pathways KGML nodes list and the RefSeq ID to Pathways Table 
```{r}
## Read KEGG KGML nodes list
full_pathway_ko_table_merged   <- read_tsv( file=file.path( kegg_ko_tables_dir, "ko_kgml_nodes_list.tsv" ))

## Read the lookup table that maps Protein RefSeq ID to KEGG Pathways 
my_hits_to_pathways_table_filt <- read_tsv( file=file.path(kegg_annotations_e_value_dir, 
                                                           "mapping_uniprot_hits_to_pathways_and_ko.tsv") ) %>%
                                  filter (  Query_Species != "Streptococcus pneumoniae")

colnames(my_hits_to_pathways_table_filt)

```

## Merge the RefSeqID to KEGG pathways table with the KGML nodes list table
```{r}

hits_to_pathways_cleaned <- my_hits_to_pathways_table_filt %>%
  # mutate( cleaned_kegg_pathway_id = str_replace( kegg_pathway_id, ".*(\\d{5}$)", "\\1")   )  %>%
  dplyr::select( Query_Species,  Query_Strain, cleaned_kegg_pathway_id, kegg_ko_id) %>%
  mutate( kegg_ko_pathway_id  = paste( "path:ko", cleaned_kegg_pathway_id, sep="" )) %>%
  dplyr::select( - cleaned_kegg_pathway_id) %>%
  distinct()

hits_to_full_pathway_map <- full_pathway_ko_table_merged %>%
  separate_rows( entry_name , sep=" ") %>%
  filter( entry_type == "ortholog") %>%
  left_join( hits_to_pathways_cleaned, by = c( "entry_name" = "kegg_ko_id",
                                               "pathway_id" = "kegg_ko_pathway_id") ) %>%
  distinct (Query_Species, Query_Strain, pathway_id, 
            pathway_title,  entry_id, kgml_type) 

```

## Calculate the percentage of protein nodes in each KEGG pathway that has at least one similar protein in the bacterial strain of interest
```{r}

# Number of nodes with at least one similar protein for each KEGG pathway and bacterial strain of interest
num_nodes_found_per_pathway_per_strain <- hits_to_full_pathway_map %>%
                                          filter( !is.na(Query_Strain)) %>%
                                          distinct() %>%
                                          group_by (Query_Species, Query_Strain, pathway_id, pathway_title, kgml_type  ) %>%
                                          summarise ( num_nodes_found =  n() ) %>%
                                          ungroup()

# The total number of KEGG ortholog (KO) group representing a protein node within each KEGG pathway
total_num_nodes_per_pathway <- full_pathway_ko_table_merged %>%
                                  filter( entry_type == "ortholog") %>%
                                group_by( pathway_id, pathway_title ) %>%
                                summarise( total_num_nodes = n()) %>%
                                ungroup()

# Calculate the percentage of nodes found per pathway and per strain
perc_nodes_found_per_pathway_per_strain <- num_nodes_found_per_pathway_per_strain %>%
  left_join( total_num_nodes_per_pathway, by = c("pathway_id", "pathway_title")) %>%
  mutate( perc_nodes_found = num_nodes_found/total_num_nodes*100 )


perc_nodes_found_per_pathway_per_strain


```

## Heat map showing the percentage of KEGG pathway nodes mapped for the metabolism KEGG pathways

The percentage of nodes that represent ortholog group of proteins in the KEGG pathways that has at least one similar sequence in the proteome of the bacterial strain of interest, with the pathways represented by the rows and bacterial strains represented by the columns.

```{r}

metabolism_perc_nodes_found_matrix <- perc_nodes_found_per_pathway_per_strain %>%
  filter ( kgml_type == "metabolism" ) %>%
  dplyr::select( -pathway_id, -kgml_type, - num_nodes_found, - total_num_nodes, -Query_Species) %>%
  # Use pivot function to convert the data from long format into wide format (closer to a matrix format) 
  pivot_wider( names_from=Query_Strain, values_from = perc_nodes_found, values_fill = list( perc_nodes_found = 0) ) %>%
  column_to_rownames( "pathway_title") %>%
  as.matrix()
  
column_annotations_metabolism <- perc_nodes_found_per_pathway_per_strain %>%
  filter ( kgml_type == "metabolism" ) %>%
  distinct ( Query_Species, Query_Strain) %>%
  dplyr::select( Query_Species ) %>%
  rename( `Query Species` = "Query_Species")


col_annot_metabolism_tsv_format <- perc_nodes_found_per_pathway_per_strain %>%
  filter ( kgml_type == "metabolism" ) %>%
  distinct ( Query_Species, Query_Strain) %>%
  mutate( column_header = paste( Query_Species, Query_Strain, sep = " " ) ) %>%
  pull( column_header )

get_rows_to_keep <- function(x){ 
  
  num_strains_required <- length(x)*proportion_of_strains_above_threshold 
  num_strains_passed <- length(which(x >= percent_kegg_nodes_mapped_threshold))
  
  return( num_strains_required <= num_strains_passed)  
}

## Show only rows with >= 30% completeness for > 80% of the strains
rows_to_include <- apply( metabolism_perc_nodes_found_matrix, 
                          1, 
                          get_rows_to_keep )

metabolism_perc_nodes_found_matrix_filt <- metabolism_perc_nodes_found_matrix[rows_to_include,]


metabolic_heatmap <- main_heatmap( metabolism_perc_nodes_found_matrix_filt, 
                                   name = "Percent",
                                   layout = list( height = 1500,
                                                  margin = list(l = 300, 
                                                               b = 100))  ) %>%
  add_col_labels() %>%
  add_row_labels() %>%
  add_col_clustering() %>%
  add_row_clustering(side = "right") %>%
  add_col_annotation(column_annotations_metabolism) 

save_iheatmap( metabolic_heatmap, 
               filename = file.path( kegg_figures_dir, 
                                     "kegg_metabolic_pathway_complete_heatmap.html"))


save_iheatmap( metabolic_heatmap, 
               filename = file.path( kegg_figures_dir, 
                                     "kegg_metabolic_pathway_complete_heatmap.png"),
               vheight = 1500 )

### test_iheatmapr_event(metabolic_heatmap, "click")

## Write the percentages as a TSV table
colnames( metabolism_perc_nodes_found_matrix) <- col_annot_metabolism_tsv_format

### Merge back with the table with data on the total number of nodes per pathway, pathway ID
metabolism_perc_nodes_found_matrix_sorted <- metabolism_perc_nodes_found_matrix %>% 
  as.data.frame %>%
  rownames_to_column(var = "KEGG_pathway_name") %>%
  left_join( total_num_nodes_per_pathway, by =c( "KEGG_pathway_name" = "pathway_title")) %>%
  rename( KEGG_pathway_id = "pathway_id",
          KEGG_pathway_total_num_nodes = "total_num_nodes") %>%
  mutate( KEGG_pathway_average_percentage = rowSums( dplyr::select(., -contains("KEGG"))    )/
                              length( vars_select(colnames(.), -contains( "KEGG")) ) ) %>%
  arrange(desc(KEGG_pathway_average_percentage)) 

### Clean up column names 
strains_column_names <- vars_select( colnames( metabolism_perc_nodes_found_matrix_sorted), -contains("KEGG"))

metabolism_perc_nodes_found_matrix_cleaned <- metabolism_perc_nodes_found_matrix_sorted %>%
  rename( pathway_id = "KEGG_pathway_id",
          total_num_nodes  = "KEGG_pathway_total_num_nodes",
          average_percent_coverage = "KEGG_pathway_average_percentage",
          pathway_name = "KEGG_pathway_name") %>%
  dplyr::select( one_of(c( "pathway_name", "pathway_id", "total_num_nodes", "average_percent_coverage", strains_column_names)) )

### Write summary statistics table
write_tsv( metabolism_perc_nodes_found_matrix_cleaned, 
           path=file.path( kegg_ko_tables_dir, "metabolism_perc_nodes_found_matrix.tsv"))

```


## Heat map showing the percentage of KEGG pathway nodes mapped for the non-metabolism KEGG pathways
The percentage of nodes that represent ortholog group of proteins in the KEGG pathways that has at least one similar sequence in the proteome of the bacterial strain of interest, with the pathways represented by the rows and bacterial strains represented by the columns.


```{r}
non_metabolism_perc_nodes_found_matrix <- perc_nodes_found_per_pathway_per_strain %>%
  filter ( kgml_type == "non-metabolism" ) %>%
  dplyr::select( -pathway_id, -kgml_type, - num_nodes_found, - total_num_nodes, -Query_Species) %>%
  # Use pivot function to convert the data from long format into wide format (closer to a matrix format) 
  pivot_wider( names_from=Query_Strain, values_from = perc_nodes_found, values_fill = list( perc_nodes_found = 0) ) %>%
  column_to_rownames( "pathway_title") %>%
  as.matrix()

column_annotations_non_metabolism <- perc_nodes_found_per_pathway_per_strain %>%
  filter ( kgml_type == "non-metabolism" ) %>%
  distinct ( Query_Species, Query_Strain) %>%
  dplyr::select( Query_Species ) %>%
  rename( `Query Species` = "Query_Species")


col_annot_non_metabolism_tsv_format <- perc_nodes_found_per_pathway_per_strain %>%
  filter ( kgml_type == "non-metabolism" ) %>%
  distinct ( Query_Species, Query_Strain) %>%
  mutate( column_header = paste( Query_Species, Query_Strain, sep = " " ) ) %>%
  pull( column_header )


## Show only rows with >= 30% completeness for > 80% of the strains


non_metabolic_rows_to_include <- apply( non_metabolism_perc_nodes_found_matrix, 
                                        1, 
                                        get_rows_to_keep )

non_metabolism_perc_nodes_found_matrix_filt <- non_metabolism_perc_nodes_found_matrix[non_metabolic_rows_to_include,]


non_metabolic_heatmap <- main_heatmap( non_metabolism_perc_nodes_found_matrix_filt, 
                                       name = "Percent",
                                       layout = list(height = 500, 
                                                     margin = list(l = 300, 
                                                                   b = 100))  ) %>%
  add_col_labels() %>%
  add_row_labels() %>%
  add_col_clustering() %>%
  add_row_clustering(side = "right") %>%
  add_col_annotation(column_annotations_non_metabolism)

save_iheatmap( non_metabolic_heatmap, 
               filename = file.path( kegg_figures_dir, 
                                     "kegg_non_metabolic_pathway_complete_heatmap.html"))

save_iheatmap( non_metabolic_heatmap, 
               filename = file.path( kegg_figures_dir, 
                                     "kegg_non_metabolic_pathway_complete_heatmap.png"),
               vheight = 35) 

### test_iheatmapr_event(non_metabolic_heatmap, "click")

## Write the percentages as a TSV table
colnames( non_metabolism_perc_nodes_found_matrix) <- col_annot_non_metabolism_tsv_format

### Merge back with the table with data on the total number of nodes per pathway, pathway ID
non_metabolism_perc_nodes_found_matrix_sorted <- non_metabolism_perc_nodes_found_matrix %>% 
  as.data.frame %>%
  rownames_to_column(var = "KEGG_pathway_name") %>%
  left_join( total_num_nodes_per_pathway, by =c( "KEGG_pathway_name" = "pathway_title")) %>%
  rename( KEGG_pathway_id = "pathway_id",
          KEGG_pathway_total_num_nodes = "total_num_nodes") %>%
  mutate( KEGG_pathway_average_percentage = rowSums( dplyr::select(., -contains("KEGG"))    )/
                              length( vars_select(colnames(.), -contains( "KEGG")) ) ) %>%
  arrange(desc(KEGG_pathway_average_percentage)) 


### Clean up column names 
strains_column_names <- vars_select( colnames( non_metabolism_perc_nodes_found_matrix_sorted), -contains("KEGG"))

non_metabolism_perc_nodes_found_matrix_cleaned <- non_metabolism_perc_nodes_found_matrix_sorted %>%
  rename( pathway_id = "KEGG_pathway_id",
          total_num_nodes  = "KEGG_pathway_total_num_nodes",
          average_percent_coverage = "KEGG_pathway_average_percentage",
          pathway_name = "KEGG_pathway_name") %>%
  dplyr::select( one_of(c( "pathway_name", "pathway_id", "total_num_nodes", "average_percent_coverage", strains_column_names)) )

### Write summary statistics table
write_tsv( non_metabolism_perc_nodes_found_matrix_cleaned, 
           path=file.path( kegg_ko_tables_dir, "non_metabolism_perc_nodes_found_matrix.tsv"))
```









