---
title: "R Notebook"
output: html_notebook
---

## Libraries management 
```{r}
if( !require(pacman)) {
  install.packages("pacman")
  library(pacman)
}

p_load(tidyverse)
p_load(iheatmapr)
p_load(tidyselect)
p_load(readxl)
p_load(magrittr)
p_load(multidplyr)
p_load(limma)

# devtools::install_github("tidyverse/multidplyr")

base_dir <- here::here() # "/home/ignatius/PostDoc/2019/Sepsis"
source( file.path( base_dir, "Source/Common/helper_functions.R") )
source( file.path( base_dir, "Source/Common/common_parameters.R") )

```


nohup time Rscript --vanilla -e "rmarkdown::render('camera_test.Rmd', output_file='camera_test.html') " > camera_test.log 2>&1 &


## Global parameter
```{r}
cluster <- new_cluster(8)
cluster_library( cluster, packages = c("tidyverse", "here", "tidyr", "magrittr"))
my_fdr_threshold <- 0.05
version_date <- master_table_version_date

results_version <- "Development"
```

## Directory Management 
```{r}
camera_test_results_dir <- file.path( owncloud_dir,  
               "Multi-omics/Results/KEGG_Enrichments/Camera_Test",  results_version) 

create_dir_if_not_exists(camera_test_results_dir)

transcriptomics_data_dir <- file.path( owncloud_dir, "Transcriptomics")
transcriptomics_master_table_dir <- file.path( owncloud_dir, "Master_Tables", version_date, "Transcriptomics" ) 
transcriptomics_abundances_dir <- file.path( transcriptomics_master_table_dir, "Abundances" ) 

```

## Files Management
```{r}

kegg_map_title_file <- file.path( owncloud_dir, 
                                     "Annotations_and_Mapping/KEGG_data_files/pathway/map_title.tab" ) 

kegg_map_title <- vroom::vroom( kegg_map_title_file, 
                                col_names = c("kegg_pathway_id", 
                                              "kegg_pathway_name") )

# To be updated 
abundances_table_file <- file.path(transcriptomics_abundances_dir, 
                                   "transcriptomics_mock_abundances_values.tsv" )


refseq_id_to_kegg_pathway_id_file <- file.path( owncloud_dir, 
                         "Annotations_and_Mapping/KEGG_Annotations_Stable_Versions", 
                         "Version_1/E_value_0.001/refseq_id_to_kegg_pathway_id_table.tsv" )


list_of_samples_file <- file.path( data_dir, "Samples_List/transcriptomics_samples_list.tsv" )


metabolite_id_to_kegg_pathway_id_file <- file.path( owncloud_dir, 
            "Master_Tables/20190807/Metabolomics/Fold_Changes/metabolomics_kegg_logFC_FDR_table.tsv" ) 


# camera_testing_output_file <- file.path( camera_test_results_dir, 
#                                                 "KEGG_camera_test_E_coli_B36_and_S_aureus_BPH2760.tsv")
# 
# camera_testing_output_rds_file <- file.path( camera_test_results_dir, 
#                                                     "KEGG_camera_test_E_coli_B36_and_S_aureus_BPH2760.RDS" )

camera_test_full_output_file <- file.path( camera_test_results_dir, 
                                                  "KEGG_camera_test_all_results.tsv")

camera_test_full_output_rds_file <- file.path( camera_test_results_dir, 
                                                      "KEGG_camera_test_all_results.RDS")

```

## Read the list of samples
```{r}

list_of_samples <- vroom::vroom( list_of_samples_file ) 

```

## Group map
# https://dplyr.tidyverse.org/reference/group_map.html

## Import KEGG annotations
```{r}

refseq_id_to_kegg_pathway_id_table <- vroom::vroom( refseq_id_to_kegg_pathway_id_file )  %>%
  dplyr::rename( Species   = "Query_Species",
          Strain    = "Query_Strain",
          entity_id = "query_acc.ver"  ) %>%
  dplyr::rename( kegg_pathway_id = "cleaned_kegg_pathway_id") %>%
  dplyr::distinct( Species, Strain, kegg_pathway_id, entity_id) %>%
  mutate( entity_type= "Protein")

metabolite_id_to_kegg_pathway_id_table_raw <- vroom::vroom( metabolite_id_to_kegg_pathway_id_file)  

metabolite_id_to_kegg_pathway_id_table <-   metabolite_id_to_kegg_pathway_id_table_raw %>%
  ## Remove loading standards
  dplyr::filter( ! ( Type_of_Experiment %in% c( "Metabolomics GC-MS", 
                                     "Metabolomics LC-MS" )  &
            str_detect( entity_id, ".*(Istd|ITSD).*" ) ) ) %>%
  dplyr::distinct( Species, Strain, entity_id, pathway_id) %>%
  dplyr::filter( !is.na(pathway_id)) %>%
  mutate( pathway_id = str_replace ( pathway_id, "^path:ko", "")) %>%
  dplyr::rename( kegg_pathway_id = "pathway_id") %>%
  dplyr::distinct( Species, Strain, kegg_pathway_id, entity_id) %>%
  mutate( entity_type= "Metabolite")
            
entity_id_to_kegg_pathway_id_table <- refseq_id_to_kegg_pathway_id_table %>%
    bind_rows( metabolite_id_to_kegg_pathway_id_table)
 
```


## Read abundances master table (draft stage)
```{r}

abundances_table <- vroom::vroom( abundances_table_file ) 

```

## Get the experimental design table 
```{r}

design_table <- abundances_table %>%
  distinct( Species, Strain, replicate_name, Treatment_Type)

```

Pooled sera
RPMI
RPMI + Glucose
RPMI + Galactose



## Create a list of gene sets, each element in the list is a list of genes contained in a gene set
```{r}
list_of_gene_sets <- entity_id_to_kegg_pathway_id_table %>%
                      group_by(Species, Strain, entity_type, kegg_pathway_id ) %>%
                      nest( gene_set = c( entity_id)) %>%
                      mutate( gene_set = purrr::map( gene_set, ~pull(., entity_id))  ) %>%
                      ungroup( Species, Strain, entity_type, kegg_pathway_id)


convert_kegg_table_to_list <- function(input_table){ 
  my_list <- input_table$gene_set 
  names(my_list) <-  input_table$kegg_pathway_id
  return(my_list)
}

list_of_gene_sets_camera <- list_of_gene_sets %>%
                            group_by( Species, Strain, entity_type )  %>%
                            nest( data = c( kegg_pathway_id, gene_set) )  %>%
                            dplyr::mutate( index = purrr::map ( data, 
                                                                convert_kegg_table_to_list   )) %>%
                            dplyr::select( -data ) %>%
                            ungroup( Species, Strain, entity_type )
                            

```

## Format the design matrix list 
```{r}

get_design_matrix <- function(input_table) {
  
  input_table <- input_table %>%
          arrange( replicate_name)
  
  Group <- factor(input_table$Treatment_Type , levels = sort(unique(input_table$Treatment_Type)))
  design_matrix<- model.matrix( ~ 0+Group )
  colnames(design_matrix) <-  c(levels(Group) )
  rownames(design_matrix ) <- input_table$replicate_name
  
  return( design_matrix)
  
}

design_table_list <- design_table %>%
    arrange( Species, Strain, replicate_name ) %>%
    dplyr::mutate( Treatment_Type = str_replace( Treatment_Type, " ", "_")%>%
                                    str_replace( "\\+", "plus")  ) %>%
    group_by( Species, Strain ) %>%
    nest( data = c (  replicate_name, Treatment_Type) )  %>%
    dplyr::mutate( design_matrix = purrr::map(data , get_design_matrix )    ) %>%
    dplyr::select(-data ) %>%
    ungroup( Species, Strain)
    
# design_table_list$design_matrix[[1]]   

```



## Format the list of abundances values for each strain and each type of comparisons
```{r}
my_pivot_norm_expr_matrix <- function( input_table) {
  
  output_matrix <- input_table %>%
    arrange( replicate_name ) %>%
    pivot_wider( names_from = replicate_name, 
                 values_from = Log_Counts) %>%
    column_to_rownames( "entity_id") %>%
    as.matrix()
  
  return( output_matrix)
  
}

list_of_abundances_table <- abundances_table %>%
  dplyr::select(-Treatment_Type, -Units ) %>%
  distinct() %>%
  group_by( Species, Strain, Type_of_Experiment ) %>%
  nest( data = c( entity_id, replicate_name, Log_Counts)) %>%
  mutate( norm_expr_matrix = purrr::map( data, my_pivot_norm_expr_matrix )   ) %>%
  dplyr::select(-data) %>%
  ungroup( Species, Strain, Type_of_Experiment )

# list_of_abundances_table$norm_expr_matrix[[1]]

  
```


## Setup the list of contrasts
```{r}

lists_of_contrasts_vectors_temp <- list ( Sera_vs_RPMI = c( 1, -1),
                                          Sera_vs_RPMI_Glucose = c( 1, 0, -1),
                                          Sera_vs_RPMI_Galactose = c( 1, -1, 0),
                                          RPMI_Galactose_vs_RPMI_Glucose = c( 0, 1, -1) )

lists_of_contrasts_vectors <- data.frame( contrast=names(lists_of_contrasts_vectors_temp)) %>%
                                mutate( contrast_vector =   lists_of_contrasts_vectors_temp  )

lists_of_contrasts_strains <- data.frame ( Species = c( "Escherichia coli",
"Klebsiella variicola",
"Klebsiella pneumoniae",
"Staphylococcus aureus",
"Streptococcus pneumoniae",
"Streptococcus pneumoniae",
"Streptococcus pneumoniae",
"Streptococcus pyogenes" ) ,
  contrast = c(  "Sera_vs_RPMI",
                 "Sera_vs_RPMI",
                 "Sera_vs_RPMI",
                 "Sera_vs_RPMI",
                 "Sera_vs_RPMI_Glucose",
                 "Sera_vs_RPMI_Galactose",
                 "RPMI_Galactose_vs_RPMI_Glucose",
                 "Sera_vs_RPMI") ) 

lists_of_contrasts <- lists_of_contrasts_strains %>%
  left_join( lists_of_contrasts_vectors, by = "contrast")



```



```{r}

camera_parameters_table <- list_of_abundances_table %>%
  mutate( entity_type = case_when(  str_detect( Type_of_Experiment, "Metabolomics") ~ 'Metabolite',
                                    TRUE ~ "Protein") ) %>%
  left_join( list_of_gene_sets_camera, 
             by = c("Species", "Strain", "entity_type"))  %>%
  left_join( design_table_list, 
             by = c("Species", "Strain") ) %>%
  left_join( lists_of_contrasts, by = "Species")


head( camera_parameters_table)
```



```{r}

camera_test_output <- camera_parameters_table %>%
                      head() %>%
                       mutate( camera_tbl =  purrr::pmap( list( y= norm_expr_matrix,
                                                                index= index, 
                                                                design = design_matrix,
                                                                contrast = contrast_vector)
                                                                , camera     ) ) 

# Error in qr.qty(QR, t(y)) : NA/NaN/Inf in foreign function call (arg 5)
# Similar problem in: 
# https://support.bioconductor.org/p/65442/
# https://github.com/davismcc/scater/issues/69

camera(camera_parameters_table$norm_expr_matrix[[1]],
       camera_parameters_table$index[[1]],
       camera_parameters_table$design_matrix[[1]],
       camera_parameters_table$contrast_vector[[1]])

# qr.qty(qr( camera_parameters_table$norm_expr_matrix[[1]]), t(camera_parameters_table$norm_expr_matrix[[1]]))


which(is.na(  camera_parameters_table$norm_expr_matrix[[1]]))

# camera_test_output$camera_tbl[[1]]

```










