---
title: "R Notebook"
output: html_notebook
author: Don Teng and Igy Pang
date: 15 October 2019
---

# Pathway Analysis for Metab Data (Sepsis)
## Do MetaboAnalystR pathway (enrichment) analysis for GC and LC datasets, separately.

### Input/Output
* Inputs: S.Aureus GC and LC strain datasets, as csv. 10 altogether: named like S_aureus_<G/L>C_<strain_num>.csv.

### Outputs:
* Lists of perturbed pathways as csvs.
* Lists of metabolites in each dataset to belong to the respective perturbed pathways, as json.

### Method Details
* Data curation for statistical analysis - Impute missing values with half of the smallest nonzero missing value. Do row-wise (sample) median normalization, then log-transform.
* Pathway enrichment - Use the globaltest algorithm. We disregard the pathway topology-based impact method and result.
* Fold change - Using the arithmetic average of log-values in each class (i.e. geometric average of non-log values), and compute fold changes for case/control.



```{r}
if( !require(pacman)) {
  install.packages("pacman")
  library(pacman)
}

p_load(here)
pacman::p_load("tidyverse", "MetaboAnalystR", "KEGGREST", "rjson")
p_load(tidyselect)
p_load(readxl)
p_load(vroom)

base_dir <- here::here() # "/home/ignatius/PostDoc/2019/Sepsis"
source( file.path( base_dir, "Source/Common/helper_functions.R") )
source( file.path( base_dir, "Source/Common/common_parameters.R") )


```


## Directory Managements
```{r}

# Get env variable: working dir

input_data_dir <- file.path(base_dir, "Data/Metabolomics/MetaboAnalystR/Data") 

output_dir <- file.path(base_dir, "Results/Metabolomics/MetaboAnalystR/Pathways_Enrichment") 

create_dir_if_not_exists(output_dir)


### ***** #### **** NEED TO Replace the temp data directory **** ### **** #### 
tempdata_dir <- output_dir <- file.path(output_dir, "TempDir") 

create_dir_if_not_exists(tempdata_dir)

```


## Files Managements
```{r}
fn_ls <- c(list.files(input_data_dir, pattern="S_aureus_GC_"), 
           list.files(input_data_dir, pattern="S_aureus_LC_"))

norm_metabolomics_data_file <- file.path( owncloud_dir, 
                "Master_Tables/20190807/Metabolomics/Heat_Map_Data", 
                "Intermediate_Files/metabolomics_log_and_normalized_values_distinct_long.tsv")

metabolites_keys_table_file <- file.path(owncloud_dir, "Metabolomics", "metabolite_keys_table.xlsx" )

```

## Read metabolites key table 
```{r}
metabolite_keys <- read_xlsx(metabolites_keys_table_file)

```


## Normalized and log transformed values 
```{r}

norm_metabolomics_data <-  vroom::vroom ( norm_metabolomics_data_file ) 

```





```{r}

call_maca_pw_analysis <- function(fn_auc_csv, kegg_species_id, perform_normalization = FALSE) {
    "Calls the pathway enrichment analysis module from MetaboAnalystR. 
    Does row-wise median-normalization and log-transforms the data.
    P-values of pathway enrichment are calculated using the `globaltest` algorithm, and pathway impact scores
    computed using the pathway centrality option. But impact should be disregarded as an overly-abstract
    graph theoretic notion that doesn't necessarily have any biological relevance. 

    PARAMS
    ------
    fn_auc_csv: str; filename of input run summary table as a csv file, with AUCs as values. 
    rownames are the sample names, column names are the metabolite names. Column 1 are the 
    experimental groupings. Because of the way this module works, only 2 groups are supported.

    RETURNS
    -------
    list of two outputs:
    tbl.out: output tibble of pathway analysis enrichment. columns:
        metabolite (compound common name), total cmpd, Hits, raw p (raw p value), -log p, 
        Holm adjust(ed p value), FDR, Impact. 
    pw.dict: named list of lists; each key is the pathway ID, and each value is a list of 
        compounds from the input data which appear in that particular pathway. 
    "

    mSet<-InitDataObjects("conc", "pathqea", FALSE)
    mSet<-Read.TextData(mSet, fn_auc_csv, "rowu", "disc");
    mSet<-CrossReferencing(mSet, "hmdb_kegg");
    mSet<-CreateMappingResultTable(mSet)
    mSet<-SanityCheckData(mSet)
    mSet<-ReplaceMin(mSet);
    mSet<-PreparePrenormData(mSet)
    if( perform_normalization == TRUE ) {
          mSet<-Normalization(mSet, "MedianNorm", "LogNorm", "NULL", ratio=FALSE, ratioNum=20)
    } else  {
          mSet<-Normalization(mSet, "NULL", "NULL", "NULL", ratio=FALSE, ratioNum=20)
    }
    mSet<-SetKEGG.PathLib(mSet, kegg_species_id)
    mSet<-SetMetabolomeFilter(mSet, F);
    mSet<-CalculateQeaScore(mSet, "rbc", "gt")

    tbl.out <- as_tibble(mSet$analSet$qea.mat, rownames="pw_name")
    pw.dict <- mSet$analSet$qea.hits

    return(list(tbl.out, pw.dict))
}


maca_pw_analysis_wrapper <- function(Species, Strain, Type_of_Experiment, data, kegg_species_id ) {
  
  my_type_of_expt <- case_when( str_detect( Type_of_Experiment, "GC-MS") ~ "GC-MS",
                                str_detect( Type_of_Experiment, "LC-MS") ~ "LC-MS",
                                TRUE ~ NA_character_ )
  
  temp_file_name <-  file.path(tempdata_dir, paste0( paste( str_replace(Species, "\\s+", "_") , 
                                               Strain, my_type_of_expt, sep="_" ),  ".csv") )

  print( temp_file_name)
  vroom::vroom_write( data, temp_file_name, delim=",", quote="all")  
  
  analysi_result <- NULL
  
    tryCatch({
 analysi_result <- call_maca_pw_analysis( temp_file_name, kegg_species_id, 
                             perform_normalization = TRUE )
  }, error=function(e){})
  
  return(analysi_result)
  
}

```


## Pivot the data matrix 
```{r}
# Sample, group, (metabolite name as remaining columns )

kegg_species_id_table <- tribble( ~Species, ~kegg_species_id,
                                  "Staphylococcus aureus", "sau",
                                  "Escherichia coli", "eco",
                                  "Streptococcus pyogenes", "spym",
                                   "Klebsiella pneumoniae", "eco",
                                  "Klebsiella variicola", "eco",
                                  "Streptococcus pneumoniae", "spn"
                                  )

cut_into_individual_tables <- norm_metabolomics_data %>%
  group_by( Species, Strain, Type_of_Experiment, sample_id, Group, entity_id  ) %>%
  summarise( log_norm_value= mean(log_norm_value) ) %>%
  ungroup() %>%
  rename( Sample = "sample_id",
          Groups = "Group") %>%
  group_by( Species, Strain, Type_of_Experiment ) %>%
  nest() %>%
  mutate( data = purrr::map(data, ~pivot_wider(., names_from = entity_id, values_from = log_norm_value))) %>%
  ungroup() %>%
  inner_join( kegg_species_id_table, by = c( "Species"))
      
# cut_into_individual_tables %>% 
#   filter(  Strain == "BPH2900" & 
#            Type_of_Experiment == "Metabolomics LC-MS" ) %>% 
#   dplyr::select(data) %>%
#   pull(data)

# cut_into_individual_tables$data

```


```{r}

maca_pw_analysis_results <- cut_into_individual_tables %>%
  dplyr::mutate( results = purrr::pmap( list(Species, Strain, Type_of_Experiment, data, kegg_species_id ),
                                        maca_pw_analysis_wrapper))

```

```{r}
# cut -f1 temp.txt | sed -e 's/\(^.*\)/"\1", /g' | perl -p -e 's/\n//'

test_kegg <- function(  kegg_species_id ) {
  
  x <- NULL 
  
  tryCatch({
 x <- call_maca_pw_analysis( "/home/ignatius/PostDoc/2019/Sepsis/Results/Metabolomics/MetaboAnalystR/Pathways_Enrichment/TempDir/Klebsiella_pneumoniae_AJ218_GC-MS.csv", kegg_species_id, 
                             perform_normalization = TRUE )
  }, error=function(e){})
   
   return(x)
  
}

list_of_ids <- c("kpu", "kpm", "kpp", "kph", "kpz", "kpv", "kpw", "kpy", "kpg", "kpc", "kpq", "kpt", 
                 "kpe", "kpo", "kpr", "kpj", "kpi", "kpa", "kps", "kpx", "kpb", "kpne", "kpnu", "kpnk", 
                 "kva", "kpk", "kvd", "kvq", "kox", "koe", "koy", "kom", "kmi", "kok", "koc", "kqu", 
                 "eae", "ear", "kqv", "kll", "klw" )

test_output <- purrr::map( list_of_ids, test_kegg)

purrr::map( test_output, 1)

```





